---
title: "BAN 3083 - Week 4: Formatting & Re-coding Variables"
author: "Student Name: [Type Your Name Here]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

if (!dir.exists("data")) dir.create("data")
if (!file.exists("data/chinook.sqlite")) {
  download.file(
    url = "https://github.com/lerocha/chinook-database/releases/download/v1.4.5/Chinook_Sqlite.sqlite",
    destfile = "data/chinook.sqlite",
    mode = "wb"
  )
}
```

## Overview: Formatting & Re-coding Variables in Business Analytics

Raw database variables are rarely ready for executive tables or statistical charts. In **BAN 3083**, formatting and re-coding variables bridges concepts you may have seen in **SQL (`CASE WHEN`)** or **SAS (`PROC FORMAT` / `IF-THEN-ELSE`)** into R:

| Data Management Task | SQL / SAS Equivalent | R (`tidyverse`) Function |
| :--- | :--- | :--- |
| Binary recoding (2 outcomes) | `IF ... THEN ... ELSE` | `if_else()` |
| Multi-tier recoding / bucketing | `CASE WHEN` / `PROC FORMAT` | `case_when()` |
| Ordering categorical levels | `PROC FORMAT` / `ORDER=DATA` | `forcats::fct_relevel()` & `fct_reorder()` |
| Cleaning & standardizing text | `UPCASE()`, `TRIM()`, `SUBSTR()` | `stringr::str_trim()`, `str_to_title()` |
| Parsing dates & day-of-week | `DATEPART()`, `WEEKDAY()` | `lubridate::ymd_hms()`, `wday()`, `month()` |

---

## Part 1: Re-coding Continuous Variables into Business Tiers (`case_when` + `forcats`)

Let's connect to `chinook.sqlite` and load `Track`, `Invoice`, and `Customer`:

```{r load-customer-spending}
library(tidyverse)
library(DBI)
library(RSQLite)
library(lubridate)
library(janitor)
library(gt)
library(scales)

con <- dbConnect(RSQLite::SQLite(), "data/chinook.sqlite")
tracks    <- dbReadTable(con, "Track")    |> as_tibble()
invoices  <- dbReadTable(con, "Invoice")  |> as_tibble()
customers <- dbReadTable(con, "Customer") |> as_tibble()
dbDisconnect(con)
```

Suppose Marketing wants every Chinook invoice classified into **Order Size Tiers**. If we just use text labels, R sorts categories **alphabetically** (putting `"Premium"` before `"Small"` and `"Standard"`!). To fix this, we format the recoded column as an **Ordered Factor** using `fct_relevel()`:

```{r recode-invoices}
invoices_recoded <- invoices |>
  mutate(
    # 1. Binary recode with if_else(): Domestic (USA) vs. International
    Market_Type = if_else(BillingCountry == "USA", "Domestic (USA)", "International"),

    # 2. Multi-level recode with case_when() (like SQL CASE WHEN / SAS PROC FORMAT)
    Order_Tier = case_when(
      Total < 3.00   ~ "Small Order (< $3)",
      Total <= 10.00 ~ "Standard Order ($3 - $10)",
      Total > 10.00  ~ "Premium Order (> $10)"
    ),

    # 3. Lock in logical business ordering (Small -> Standard -> Premium)
    Order_Tier = fct_relevel(
      Order_Tier,
      "Small Order (< $3)",
      "Standard Order ($3 - $10)",
      "Premium Order (> $10)"
    )
  )
```

### Displaying Recoded Variables in a Cross-Tabulated Table (`tabyl`)
Let's inspect the distribution of our newly recoded `Order_Tier` across `Market_Type`:

```{r tabyl-order-tiers}
invoices_recoded |>
  tabyl(Order_Tier, Market_Type) |>
  adorn_totals(c("row", "col")) |>
  adorn_percentages("col") |>
  adorn_pct_formatting(digits = 1) |>
  adorn_ns() |>
  gt() |>
  tab_header(
    title = "Invoice Order Size Tiers: Domestic vs. International",
    subtitle = "Recoded via if_else(), case_when(), and ordered via fct_relevel()"
  )
```

### Plotting Recoded Factors (`ggplot2`)
Notice on the x-axis below how `Order_Tier` appears in **exact business order** (`Small` $\rightarrow$ `Standard` $\rightarrow$ `Premium`) rather than alphabetical order!

```{r plot-order-tiers}
ggplot(invoices_recoded, aes(x = Order_Tier, fill = Market_Type)) +
  geom_bar(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Domestic (USA)" = "#1b9e77", "International" = "#d95f02")) +
  labs(
    title = "Distribution of Chinook Invoices by Order Size Tier",
    subtitle = "Factor levels ordered logically with forcats::fct_relevel()",
    x = "Recoded Order Size Tier",
    y = "Number of Invoices",
    fill = "Market Segment"
  ) +
  theme_minimal(base_size = 12)
```

---

## Part 2: String Formatting & Text Standardization (`stringr`)

Real-world customer databases often have inconsistent capitalization, extra spaces, or email domains we want to extract for B2B vs. B2C segmentation. Let's use `stringr` functions on the Chinook `customers` table:

```{r stringr-cleaning}
customers_formatted <- customers |>
  mutate(
    # Combine and format full name in Title Case
    Full_Name = str_to_title(paste(str_trim(FirstName), str_trim(LastName))),

    # Extract the email provider domain (everything after the '@' symbol)
    Email_Domain = str_extract(Email, "(?<=@).+"),

    # Recode Corporate vs. Consumer Customer based on whether Company is missing
    Account_Category = if_else(is.na(Company) | Company == "", "Individual Consumer (B2C)", "Corporate Account (B2B)")
  )

# Display the B2B/B2C breakdown in a formatted table
customers_formatted |>
  count(Account_Category, sort = TRUE) |>
  mutate(Share = n / sum(n)) |>
  gt() |>
  tab_header(title = "Chinook Customer Base: Corporate (B2B) vs. Consumer (B2C)") |>
  fmt_percent(columns = Share, decimals = 1)
```

---

## Part 3: Date Formatting & Seasonality Extraction (`lubridate`)

In SQLite, `InvoiceDate` is stored as a plain text string (`"2009-01-01 00:00:00"`). Using **`lubridate`**, we can parse text strings into true Date objects and extract the **Year**, **Month Name**, and **Quarter** to visualize sales seasonality:

```{r lubridate-formatting}
invoices_dated <- invoices_recoded |>
  mutate(
    # Parse text timestamp into a Date-Time object
    Invoice_Date = ymd_hms(InvoiceDate),
    Invoice_Year = year(Invoice_Date),
    # Extract ordered Month abbreviation (Jan, Feb, Mar, ..., Dec)
    Invoice_Month = month(Invoice_Date, label = TRUE, abbr = TRUE),
    # Extract Fiscal Quarter (Q1, Q2, Q3, Q4)
    Fiscal_Quarter = paste0("Q", quarter(Invoice_Date))
  )

# Summarize revenue by Month and Fiscal Quarter
monthly_seasonality <- invoices_dated |>
  group_by(Invoice_Month, Fiscal_Quarter) |>
  summarize(
    Invoice_Count = n(),
    Total_Revenue = sum(Total),
    .groups = "drop"
  )
```

### Displaying Monthly Seasonality with a Plot (`ggplot2`)

```{r plot-monthly-seasonality}
ggplot(monthly_seasonality, aes(x = Invoice_Month, y = Total_Revenue, fill = Fiscal_Quarter)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = label_dollar()) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Chinook Cumulative Revenue by Calendar Month",
    subtitle = "Month labels & quarters extracted from raw timestamps via lubridate::month() and quarter()",
    x = "Month of Year",
    y = "Total Revenue (USD)",
    fill = "Fiscal Quarter"
  ) +
  theme_minimal(base_size = 12)
```

---

## Part 4: Your Turn! Recoding & Visualizing Track Lengths

Run and inspect the code below, which converts `Milliseconds` into `Minutes` in the Chinook `Track` table, recodes tracks into **4 Song Length Categories**, displays a formatted `gt` summary table, and plots the file size breakdown!

```{r exercise-track-recode}
tracks_recoded <- tracks |>
  mutate(
    Minutes = Milliseconds / 60000,
    Megabytes = Bytes / (1024^2),
    Length_Category = case_when(
      Minutes < 2.5  ~ "1. Short Track (< 2.5 min)",
      Minutes <= 4.5 ~ "2. Standard Radio (2.5 - 4.5 min)",
      Minutes <= 8.0 ~ "3. Extended Track (4.5 - 8 min)",
      TRUE           ~ "4. Long-Form / Video (> 8 min)"
    )
  )

# 1. Display Formatted Summary Table
tracks_recoded |>
  group_by(Length_Category) |>
  summarize(
    Track_Count = n(),
    Avg_Minutes = mean(Minutes),
    Avg_Megabytes = mean(Megabytes),
    Avg_Price = mean(UnitPrice)
  ) |>
  gt() |>
  tab_header(title = "Chinook Catalog Breakdown by Recoded Song Duration") |>
  fmt_number(columns = c(Avg_Minutes, Avg_Megabytes), decimals = 2) |>
  fmt_currency(columns = Avg_Price, currency = "USD")
```

```{r exercise-track-plot}
# 2. Display Boxplot of File Size (MB) across Recoded Length Categories
tracks_recoded |>
  filter(Megabytes < 50) |>
  ggplot(aes(x = Length_Category, y = Megabytes, fill = Length_Category)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  coord_flip() +
  scale_fill_viridis_d(option = "mako", guide = "none") +
  labs(
    title = "File Size (MB) Distribution Across Recoded Track Length Categories",
    x = "Recoded Track Duration Tier",
    y = "File Size (Megabytes)"
  ) +
  theme_minimal(base_size = 12)
```

---

## 🌐 How to Publish to RPubs & Submit Your URL on Canvas

1. At the top of this file (Line 3), replace `[Type Your Name Here]` with your name.
2. Click **🧶 Knit**, then click the blue **Publish** button $\rightarrow$ **RPubs**.
3. Submit your `https://rpubs.com/...` URL on **Canvas**!

```{r rpubs-backup, eval=FALSE}
rsconnect::rpubsUpload(
  title = "BAN 3083: Week 4 Formatting & Recoding Variables",
  contentFile = "Week04_Formatting_and_Recoding.html",
  originalDoc = "Week04_Formatting_and_Recoding.Rmd"
)$continueUrl
```
