---
title: "Week 1: From SQL to R with the Chinook Database"
author: "Student Name: [Type Your Name Here]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Safety check: if running outside Binder without chinook.sqlite, download it automatically
if (!dir.exists("data")) dir.create("data")
if (!file.exists("data/chinook.sqlite")) {
  download.file(
    url = "https://github.com/lerocha/chinook-database/releases/download/v1.4.5/Chinook_Sqlite.sqlite",
    destfile = "data/chinook.sqlite",
    mode = "wb"
  )
}
```

## Welcome to RStudio! (Meet the 4 Panes)

Take a quick look around your browser window—you are looking at the **4-pane RStudio IDE**:

1. **Top-Left (Source Editor — *where you are reading this!*)**: Your notebook and scripts. Click the green **▶ (Run Current Chunk)** button on the top-right of any gray code box below to run code.
2. **Bottom-Left (Console)**: Where R executes commands behind the scenes.
3. **Top-Right (Environment)**: Shows all the datasets and variables currently loaded in memory.
4. **Bottom-Right (Files / Plots / Help)**: Where your visualizations (`Plots`) and files (`Files`) appear.

---

## Part 1: Connecting R to the Chinook SQLite Database

In our SQL module, you queried the **Chinook digital music store database** (`Artist`, `Album`, `Track`, `Genre`, `Customer`, `Invoice`). 

R doesn't replace SQL—in fact, R works *hand-in-hand* with SQL! Let's load our packages and connect to `chinook.sqlite`.

👉 **Click the green `▶` play button on the right side of the gray box below:**

```{r connect-db}
library(DBI)
library(RSQLite)
library(tidyverse)
library(dbplyr)

# Connect to the Chinook SQLite file
con <- dbConnect(RSQLite::SQLite(), "data/chinook.sqlite")

# List all tables in the database (these should look familiar!)
dbListTables(con)
```

---

## Part 2: Writing Pure SQL Directly Inside RMarkdown

Did you know you can write **100% pure SQL** inside an RMarkdown document without learning a single line of R syntax first?

Look at the header of the code chunk below: `{sql connection=con, output.var="genre_summary"}`.
- `connection=con` tells RStudio to run this SQL query against our Chinook database.
- `output.var="genre_summary"` tells RStudio to save the resulting table as a new R dataset named `genre_summary`!

👉 **Run the SQL chunk below, then look at your Top-Right (Environment) pane:**

```{sql connection=con, output.var="genre_summary"}
SELECT 
    g.Name AS Genre,
    COUNT(t.TrackId) AS TrackCount,
    ROUND(AVG(t.Milliseconds) / 60000.0, 2) AS AvgMinutes,
    ROUND(AVG(t.UnitPrice), 2) AS AvgPrice
FROM Track AS t
INNER JOIN Genre AS g ON t.GenreId = g.GenreId
GROUP BY g.Name
HAVING COUNT(t.TrackId) >= 30
ORDER BY TrackCount DESC;
```

Let's inspect the `genre_summary` table that SQL just handed over to R:

```{r view-genre-summary}
genre_summary
```

---

## Part 3: Running SQL Inside R Code with `dbGetQuery()`

Another common way data scientists run SQL in R is using the `dbGetQuery()` function. You pass it your database connection (`con`) and a SQL query string, and it returns an R data frame.

Let's find the **Top 10 Countries by Total Invoice Sales** in Chinook:

```{r sql-in-r}
country_sales <- dbGetQuery(con, "
  SELECT 
      BillingCountry AS Country,
      COUNT(InvoiceId) AS NumInvoices,
      ROUND(SUM(Total), 2) AS TotalRevenue
  FROM Invoice
  GROUP BY BillingCountry
  ORDER BY TotalRevenue DESC
  LIMIT 10;
")

country_sales
```

---

## Part 4: Why Combine SQL and R? Visualization (`ggplot2`)!

SQL is fantastic at joining and aggregating tables—but SQL cannot create graphics. 

Because our SQL query results (`genre_summary` and `country_sales`) are now R data frames, we can pipe them directly into **`ggplot2`**, R's industry-standard visualization library!

👉 **Run the chunk below to visualize the SQL query we just ran:**

```{r plot-genre}
ggplot(genre_summary, aes(x = reorder(Genre, TrackCount), y = TrackCount, fill = AvgMinutes)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma", name = "Avg Track\nLength (min)") +
  labs(
    title = "Chinook Catalog: Most Popular Genres & Track Lengths",
    subtitle = "Data extracted via SQL JOIN on Track and Genre tables",
    x = "Genre",
    y = "Number of Tracks (n >= 30)"
  ) +
  theme_minimal(base_size = 12)
```

---

## Part 5: Sneak Peek at Next Week — How `dplyr` Speaks SQL!

Starting next week, we will learn **`dplyr`**, R's grammar of data wrangling. 

Here is a secret that makes learning `dplyr` much easier: **`dplyr` verbs are really just SQL clauses in disguise!**

| What you want to do | SQL Clause | R `dplyr` Verb |
| :--- | :--- | :--- |
| Pick columns | `SELECT` | `select()` |
| Filter rows | `WHERE` / `HAVING` | `filter()` |
| Create new columns | `SELECT ... AS` | `mutate()` |
| Group rows | `GROUP BY` | `group_by()` |
| Aggregate metrics | `COUNT()`, `AVG()`, `SUM()` | `summarize()` |
| Sort rows | `ORDER BY` | `arrange()` |

In fact, R can connect directly to a SQL table using `tbl(con, "TableName")` and **automatically translate your R code into SQL**!

👉 **Run the chunk below to see R write the SQL query for you using `show_query()`:**

```{r dplyr-to-sql}
# Connect to the Track and Genre tables lazily
tracks_db <- tbl(con, "Track")
genres_db <- tbl(con, "Genre")

# Write R (dplyr) code using the pipe operator |>
r_pipeline <- tracks_db |>
  inner_join(genres_db, by = "GenreId") |>
  group_by(Genre = Name.y) |>
  summarize(
    TrackCount = count(),
    AvgMinutes = mean(Milliseconds, na.rm = TRUE) / 60000.0
  ) |>
  filter(TrackCount >= 30) |>
  arrange(desc(TrackCount))

# Ask R: "What SQL query did you generate behind the scenes?"
show_query(r_pipeline)
```

And if we execute `collect(r_pipeline)`, R runs that exact query and brings back the results:

```{r collect-results}
collect(r_pipeline)
```

---

## Part 6: Your Turn! (Lab Exercises)

### Exercise 1: Write Your Own SQL Query in R
Modify the `{sql}` chunk below to find the **Top 10 Artists with the most Albums** in the Chinook database (joining `Artist` and `Album`). The result will automatically be saved to R as `top_artists`.

```{sql connection=con, output.var="top_artists"}
SELECT 
    ar.Name AS ArtistName,
    COUNT(al.AlbumId) AS AlbumCount
FROM Artist AS ar
INNER JOIN Album AS al ON ar.ArtistId = al.ArtistId
GROUP BY ar.ArtistId, ar.Name
ORDER BY AlbumCount DESC
LIMIT 10;
```

### Exercise 2: Visualize Your SQL Output with `ggplot2`
Run the chunk below to plot your `top_artists` SQL result. Try changing `fill = "steelblue"` to another color (like `"darkorange"`, `"seagreen"`, or `"firebrick"`) and updating the chart `title`!

```{r plot-top-artists}
ggplot(top_artists, aes(x = reorder(ArtistName, AlbumCount), y = AlbumCount)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Artists by Number of Albums in Chinook",
    x = "Artist",
    y = "Number of Albums"
  ) +
  theme_minimal()
```

```{r cleanup, include=FALSE}
# Always good practice to disconnect when knitting finishes
dbDisconnect(con)
```

---

## 🌐 How to Publish to RPubs & Submit Your URL on Canvas

Instead of downloading files, you will publish your finished report to **[RPubs.com](https://rpubs.com/)** and submit your public web link on Canvas!

### Step 1: Knit & Publish
1. At the top of this file (Line 3), replace `[Type Your Name Here]` with your name.
2. Click the **🧶 Knit** button at the top of this pane. Your formatted HTML report will open in a preview window.
3. In the top-right corner of that preview window (or the top-right of this editor pane), click the blue **Publish** button (the blue eye/circle icon) $\rightarrow$ choose **RPubs**.
4. A browser tab will open on **RPubs.com**:
   - Sign in (or create a free RPubs account if this is your first lab).
   - Give your report a title (e.g., `Stats 101 - Week 1 Chinook Lab`) and click **Continue**.
5. Copy the URL from your browser address bar (it will look like `https://rpubs.com/your_username/1234567`) and **paste that URL into the Canvas assignment submission box!**

> 💡 **Popup Blocker Backup:** If your browser blocks the Publish popup window, just run the code chunk below after clicking **Knit**—it will print a direct clickable link in the Console to finish publishing your report to RPubs!

```{r rpubs-backup, eval=FALSE}
# Only run this if the blue "Publish -> RPubs" button was blocked by your browser:
upload_result <- rsconnect::rpubsUpload(
  title = "Stats 101: Week 1 SQL to R (Chinook)",
  contentFile = "Week01_SQL_to_R_Chinook.html",
  originalDoc = "Week01_SQL_to_R_Chinook.Rmd"
)
upload_result$continueUrl
```
