---
title: "Week 1: From SQL to R with the Chinook Database"
author: "Student Name: [Type Your Name Here]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Safety check: if running outside Binder without chinook.sqlite, download it automatically
if (!dir.exists("data")) dir.create("data")
if (!file.exists("data/chinook.sqlite")) {
  download.file(
    url = "https://github.com/lerocha/chinook-database/releases/download/v1.4.5/Chinook_Sqlite.sqlite",
    destfile = "data/chinook.sqlite",
    mode = "wb"
  )
}
```
## Welcome to RStudio! (Meet the 4 Panes)
Take a quick look around your browser window—you are looking at the **4-pane RStudio IDE**:
1. **Top-Left (Source Editor — *where you are reading this!*)**: Your notebook and scripts. Click the green **▶ (Run Current Chunk)** button on the top-right of any gray code box below to run code.
2. **Bottom-Left (Console)**: Where R executes commands behind the scenes.
3. **Top-Right (Environment)**: Shows all the datasets and variables currently loaded in memory.
4. **Bottom-Right (Files / Plots / Help)**: Where your visualizations (`Plots`) and files (`Files`) appear.
---
## Part 1: Connecting R to the Chinook SQLite Database
In our SQL module, you queried the **Chinook digital music store database** (`Artist`, `Album`, `Track`, `Genre`, `Customer`, `Invoice`). 
R doesn't replace SQL—in fact, R works *hand-in-hand* with SQL! Let's load our packages and connect to `chinook.sqlite`.
👉 **Click the green `▶` play button on the right side of the gray box below:**
```{r connect-db}
library(DBI)
library(RSQLite)
library(tidyverse)
library(dbplyr)
# Connect to the Chinook SQLite file
con <- dbConnect(RSQLite::SQLite(), "data/chinook.sqlite")
# List all tables in the database (these should look familiar!)
dbListTables(con)
```
---
## Part 2: Writing Pure SQL Directly Inside RMarkdown
Did you know you can write **100% pure SQL** inside an RMarkdown document without learning a single line of R syntax first?
Look at the header of the code chunk below: `{sql connection=con, output.var="genre_summary"}`.
- `connection=con` tells RStudio to run this SQL query against our Chinook database.
- `output.var="genre_summary"` tells RStudio to save the resulting table as a new R dataset named `genre_summary`!
👉 **Run the SQL chunk below, then look at your Top-Right (Environment) pane:**
```{sql connection=con, output.var="genre_summary"}
SELECT 
    g.Name AS Genre,
    COUNT(t.TrackId) AS TrackCount,
    ROUND(AVG(t.Milliseconds) / 60000.0, 2) AS AvgMinutes,
    ROUND(AVG(t.UnitPrice), 2) AS AvgPrice
FROM Track AS t
INNER JOIN Genre AS g ON t.GenreId = g.GenreId
GROUP BY g.Name
HAVING COUNT(t.TrackId) >= 30
ORDER BY TrackCount DESC;
```
Let's inspect the `genre_summary` table that SQL just handed over to R:
```{r view-genre-summary}
